Probe or Wait: Handling tail losses using Multipath TCP

### 2017-03-08 main.tex
Update

Multipath TCP(MPTCP) is an experimental standard proposed as an extension to TCP~\cite{rfc6824}. It allows devices with multiple interfaces to transmit data simultaneously on both interfaces, there by improving the throughput of end to end connection. Using MPTCP, a connection can have multiple TCP subflows using different interfaces on different routes. Each TCP subflow imparts the functional behavior of a standard TCP flow.

TCP has mainly two mechanisms for detecting and recovering from packet losses namely Fast Retransmit (FR) and Retransmission Timeout~(RTO)~\cite{Flach:2013}. Fast retransmit is triggered on receipt of a predecided number of duplicate acknowledgements, considering it as a loss indication. In case of RTO, sender waits for the packet loss to trigger a timeout before retransmitting that packet. For long flows where there are sufficiently large number of packets to be transferred, FR is quicker than RTO in detecting and recovering packet losses. On the other hand, shorter flows have to depend on RTO for packet loss detection and recovery, thus incurring more latency. A single packet loss in a short flow may take many RTTs to detect and recover. This scenario is also applicable to the packets at the end of the flow (or tail) in a long flow. Though, FR and RTO are reliable even in tail loss recovery, they are not efficient in terms of recovery completion time.

 MPTCP connection starts with one TCP flow to send data and with available network interfaces and protocol support, additional flows are added. MPTCP uses a connection level congestion window as well as subflow level congestion windows for each subflow. Data is split across the flows and each data packet is scheduled on one of the subflows based on minimum RTT. Packet losses in each subflow are detected and recovered in a similar fashion as that of TCP. However, the MPTCP standard does not restrict on, how the loss recovery should happen in the implementation and which subflow should be retransmit the lost packets. Loss recovery can be conducted at different levels: the meta-level and the subflow level. If the recovery is handled at the meta level, the lost packet may be rescheduled and retransmitted at the available subflow with lowest RTT. If the recovery is handled at the flow level, the packet may be retransmitted in the same subflow. An important part of MPTCP packet loss recovery process is to follow the TCP semantics and must retransmit lost packet on the initial path of transfer.

TCP Loss Probe (TLP)~\cite{Flach:2013} is a mechanism to detect and recover from tail losses by employing a loss probe packet for a timeout shorter than the RTO. Probe timeout enables the short flows to recover packet loss at the tail by triggering the fast retransmit. It assumes other algorithms such as Early Retransmit~(ER)\cite{rfc5827} and Forward Acknowledgement~(FACK)~\cite{FACK} threshold based recovery are enabled. TLP is available in Linux kernel and enabled by default. Further evaluations in~\cite{Rajiullah:2015}, showed that TLP provides significant reductions in latency for short flows with one to n-degree tail losses.

In this paper, we analyze the retransmission behavior of MPTCP in selected tail loss cases to identify the scope for improvement. We propose a change in the retransmission strategy of MPTCP to improve the latency performance. We provide the related research on MPTCP retransmission in section~\ref{relwork}, and experimental setup in section~\ref{exsetup}. We discuss the observations from the experiments in section~\ref{disc}, that lead to our proposed improvement on current Linux implementation of TLP for MPTCP in section~\ref{impr}. In section~\ref{eval}, we provide an evaluation of the proposed TLP for MPTCP implementation in Linux and conclude the paper in section~\ref{conc}.

@@

Retransmission schemes of TCP are of significant importance in achieving lower flow completion time. Several research studies led to improved retransmission strategies for TCP. The current Linux TCP implementation provides space for improvement in specific cases of packet loss scenarios~\cite{prr}. In a TCP flow, sequence numbers are used to identify the lost packets and retransmit the respective packets. Multipath TCP uses two levels of sequence numbers to support efficient data transfer, namely data sequence and TCP sequence. Data sequence is for the end to end data transfer and TCP sequence number is for individual flow data sequence similar to TCP. A loss occurring in an individual TCP flow corresponds to a loss in end to end data. Multipath TCP, as an extension of TCP, uses the basics of TCP retransmission. However, the interaction with two levels makes the problem of retransmission more challenging than in TCP. The dual sequence numbering enables MPTCP to respond to loss of packets by retransmitting the lost packets on an alternate path. It allows a packet to be retransmitted on a different subflow than that of the originally sent flow.

The Linux implementation of MPTCP uses a set of retransmission heuristics to decide on the retransmissions. The data outstanding on a timed out subflow should be rescheduled for transmission on a different subflow using timeout as the indicator. Fast retransmit on a subflow will not trigger retransmission on another subflow. The target devices of MPTCP contain a degree of asymmetry in their interface characteristics such as 3G or 4G vs WLAN, leading to flows with different delay characteristics. Authors in~\cite{fuso}, try to exploit the path diversity by quickly retransmitting on the fast paths. Quick retransmission comes with a cost of redundant packets as each individual TCP flow should retransmit always to follow TCP semantics. MPTCP standard suggests more conservative approaches. MPTCP is designed to exploit the multiple interfaces and often one or more of these interfaces could be wireless with completely different characteristics than that of wired interfaces. Authors in~\cite{Shin} argue that the calculation RTO for MPTCP flows should include the interface characteristics to improve loss recovery time.

@@

This paper analyzes the performance of state of the art loss recovery mechanism of MPTCP and presents a case for improvement in the tail loss scenario. 
For understanding the loss recovery and retransmission policies in MPTCP Linux implementation version 0.91, we consider several specific cases of tail losses.  
No tuning of TCP settings such as, adjusting the delayed ACK behavior are performed. We consider the standard settings of the Linux kernel. We consider mainly various cases of tail loss and its recovery.
The Linux implementation of TCP is considered as the baseline TCP retransmission for implementation flexibility and availability of TLP support.

@@

Experiments use the CORE emulation platform~\cite{CORE} with a simple topology as depicted in Figure~\ref{fig1}.
with client node connected to two wireless interfaces 3G or 4G and WLAN and server node connected to a wired router.
Characteristics of the connection setup and assumptions about the parameters are provided in Table~\ref{tab1}.
The parameters are chosen to be close to real values of bandwidth and delay of respective technologies.
There is no attempt in this study to focus on the effect of link detailed characteristics in retransmission performance.

@@

\subsection{Testing retransmission with deterministic loss patterns}
In order to understand the retransmission behavior of the Linux MPTCP implementation and to reproduce the observed retransmissions, we use a deterministic drop pattern.
Losses are generated by associating netem with corresponding interfaces and dropping select packets. This process is simplified by using the KAUNetem tool~\cite{Garcia2016}. 
MPTCP connection starts with a single TCP subflow and subsequently, one or more subflows are added following an agreement between client and server on the MPTCP protocol 
support and interface availability. So one has to wait in time and packets for the second subflow establishment to understand the MPTCP retransmissions in multiple flows. We send two bursts of 80 
packets each and drop tail packets on one of the interfaces. This is to ensure that the necessary and sufficient conditions for probe triggering are met and tail loss probe is generated on that
subflow. Three test cases are evaluated with one packet tail loss, two packet tail loss and one packet along with probe~(in TLP) or retransmission~(in RTO) loss.

>>>>

Multipath TCP~(MPTCP) is an experimental standard proposed as an extension to TCP~\cite{rfc6824}. It allows devices with multiple interfaces to transmit data simultaneously on both interfaces, thereby improving the throughput of end-to-end connections. Using MPTCP, a connection can have multiple TCP subflows using different interfaces on different routes. Each TCP subflow imparts the functional behavior of a standard TCP flow.

TCP has two mechanisms for detecting and recovering from packet losses: Fast Retransmit (FR) and Retransmission Timeout~(RTO)~\cite{Flach:2013}. Fast retransmit is triggered on receipt of a predecided number of duplicate acknowledgements, considering it as a loss indication. In case of RTO, a sender waits for the packet loss to trigger a timeout before retransmitting that packet. During timeout, the sender cannot send any packets, but with FR the out-of-order packets are presumed lost upon receiving duplicate acknowledgements. The sender can start retransmission without waiting for timeout. For flows where there are sufficiently many packets to be transferred, FR is quicker than RTO in detecting and recovering packet losses. Fast retransmit coupled with fast recovery can send data outstanding whene there is enough data to send in long flows. On the other hand, shorter flows have to depend on RTO for packet loss detection and recovery, thus incurring more latency. A single packet loss in a short flow may take many RTTs to detect and recover. This scenario is also applicable to the packets at the end of the flow (or tail), in a long flow. Although, RTO is reliable, even in tail loss recovery, it is not efficient in terms of recovery completion time.

 MPTCP connections start with one TCP flow to send data over, and may add more flows to the connection if configured to. MPTCP uses a connection level congestion window as well as subflow level congestion windows for each subflow. Data is split across the flows and each data packet is scheduled on one of the subflows based on minimum RTT. Packet losses in each subflow are detected and recovered in a similar fashion as that of TCP. Loss recovery can be conducted at different levels: the MPTCP connection level (meta level) and at subflow level. If the recovery is handled at meta level, a lost packet may be rescheduled and retransmitted on the available subflow with the lowest RTT. If recovery is handled at the subflow level, the packet may be retransmitted on the same subflow. An important part of MPTCP packet loss recovery process is to follow the TCP semantics and retransmit lost packets on the initial path of transfer. However, the MPTCP standard does not restrict on, how the loss recovery should happen in the implementation and which subflow should retransmit the lost packets.

TCP Loss Probe (TLP)~\cite{Flach:2013} is a mechanism to detect and recover from tail losses by employing a loss probe packet for a timeout shorter than the RTO. Probe timeout enables short flows to recover packet loss at the tail, by triggering FR. It assumes other algorithms such as Early Retransmit~(ER)\cite{rfc5827} and Forward Acknowledgement~(FACK)~\cite{FACK} threshold based recovery are enabled. TLP is available in Linux kernel and enabled by default. Further evaluations in~\cite{Rajiullah:2015}, showed that TLP provides significant reductions in latency for short flows with one to n-degree tail losses.

In this paper, we analyze the retransmission behavior of MPTCP in selected tail loss cases to identify the scope for improvement. We propose a change in the retransmission strategy of MPTCP to improve the latency performance. We provide related research on MPTCP retransmission in section~\ref{relwork}, and the experimental setup in section~\ref{exsetup}. We discuss observations from the experiments in section~\ref{disc}. Based on the observations, we propose an improvement to the current Linux implementation of TLP for MPTCP in section~\ref{impr}. In section~\ref{eval}, we provide an evaluation of the proposed TLP for MPTCP mechanism and conclude the paper in section~\ref{conc}.

@@

Retransmission schemes are of significant importance in achieving low flow completion time. Several research studies has led to improved retransmission strategies for TCP. The current Linux TCP implementation provides room for improvement in several packet loss scenarios.

In a TCP flow, sequence numbers are used to identify lost packets and retransmit them. MPTCP uses two levels of sequence numbers to support efficient data transfer, namely data sequence and TCP sequence. Data sequence is for the end-to-end data transfer and TCP sequence number is for individual flow data sequence similar to TCP. There is an association between data sequence number and TCP sequence number within a subflow.  A loss occurring in an individual TCP flow corresponds to a loss in end-to-end data. Multipath TCP, as an extension of TCP, uses the basics of TCP retransmission. However, the interaction with two levels makes the problem of retransmission more challenging than in TCP. The dual sequence numbering enables MPTCP to respond to loss of packets by retransmitting the lost packets on an alternate path.

The Linux implementation of MPTCP uses a set of retransmission heuristics to handle retransmissions. The data outstanding on a timed out subflow should be rescheduled for transmission on a different subflow using timeout as the indicator. Fast retransmit on a subflow will not trigger retransmission on another subflow. The target devices of MPTCP contain a degree of asymmetry in their interface characteristics such as 3G or 4G vs WLAN, leading to flows with different delay characteristics. Authors in~\cite{fuso}, try to exploit the path diversity by quickly retransmitting on the fastest paths. Quick retransmission comes with a cost of redundant packets as each individual TCP flow should retransmit always to follow TCP semantics. MPTCP standard suggests a more conservative approach. MPTCP is designed to exploit the multiple interfaces and often one or more of these could be wireless with completely different characteristics than that of wired interfaces. Authors in~\cite{Shin} argue that the calculation RTO for MPTCP flows should include the interface characteristics to improve loss recovery time.

@@

This paper analyzes the performance of state of the art loss recovery mechanism of MPTCP and presents a case for improvement in the tail loss scenario. 
For understanding the loss recovery and retransmission policies in MPTCP Linux implementation version 0.91, we consider several specific cases of tail losses.  
We use the standard settings of the Linux kernel to understand the performance for various cases of tail loss and its recovery.
The Linux implementation of TCP is considered as the baseline TCP retransmission for implementation flexibility and availability of TLP support.

@@

Experiments use the CORE emulation platform~\cite{CORE} with a simple topology as depicted in Figure~\ref{fig1}
with client node connected to two wireless interfaces 3G or 4G and WLAN and server node connected to a wired router.
CORE emulates the Linux networking stack in each node of the topology by using lightweight virtualization technique known as Linux network namespaces.
Characteristics of the connection setup and assumptions about the parameters are provided in Table~\ref{tab1}.
The parameters are chosen to be close to real values of bandwidth and delay of respective technologies.
There is no attempt in this study to focus on the effect of link detailed characteristics in retransmission performance.

@@

\subsection{Testing retransmission with deterministic loss patterns}
In order to understand the retransmission behavior of the Linux MPTCP implementation and to reproduce the observed retransmissions, we use a deterministic drop pattern.
Losses are generated by associating netem with corresponding interfaces and dropping select packets. This process is simplified by using the KAUNetem tool~\cite{Garcia2016}, which allows experimenter to e.g., drop packets based on their positions in a flow. MPTCP connection starts with a single TCP subflow and subsequently, one or more subflows are added following an agreement between the client and the server. So one has to wait both in time and packets for the second subflow establishment. We send two bursts of 80 packets each and drop tail packets at one of the interfaces. This is to ensure that the necessary and sufficient conditions for probe triggering are met and tail loss probe is generated on that
subflow. Three test cases are evaluated with one packet tail loss, two packet tail loss and one packet along with probe~(in TLP) or retransmission~(in RTO) loss.
